Software Architectures to Integrate Workflow Engines in Science Gateways

### 2016-04-28 paper.tex
Replaced 'global complexity' with 'integration challenge'

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\closedanswerednote}[6]{\pdfmargincomment[color=gray,author=#1-#3-#5,open=true]{#1: #2 \textCR \textCR #3: #4 \textCR \textCR #5: #6}}

@@

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. We abstract software
  components and interactions in 6 architectures that we use to
  classify 18 systems. We compare these architectures using metrics
  measuring global complexity, robustness, extensibility, scalability
  and other specific features. Results show the pros and cons of the
  different architectures: tight integration and task encapsulation are the
  simplest architectures and should therefore be the easiest to
  implement and the most robust ones, most architectures perform the
  same in terms of extensibility, the pool model is the most scalable
  and meta-workflows are only available in nested workflows and
  workflow import. Our analysis provides insights for software
  architects on how to integrate science gateways and workflow
  engines.
\end{abstract}

@@

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework based on the functional
interactions between the main software components involved. In
Section~\ref{sec:evaluation}, we introduce metrics to measure
global complexity, robustness, extensibility, scalability and other
specific features of the architectures. We use these metrics to
evaluate and compare the 6 architectures. The paper closes on a
discussion and conclusion that highlight architectural recommendations
for science gateway and workflow engine architects.

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: global complexity, execution complexity,
extensibility, scalability and other specific features. Criteria break
down to specific metrics where \emph{lower value indicates better
  performance}. For each criterion, a total score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that they can be
summed. The criteria and metrics are explained hereafter.

@@

\cellcolor[HTML]{EEEEEE}\textbf{Global complexity}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\

@@

\emph{Global complexity} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain the
total score for global complexity. Global complexity measures the
total number of software pieces to develop, including interactions and
components. It may be used to measure the development effort required
to build the architecture.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity since complex architectures tend to be more prone to
failure. More precisely, robustness is determined as the number of
software components and interactions that are specific to workflow
execution:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are the ones represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the total score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\paragraph{Global complexity} This architecture does not require any
component in addition to the science gateway and infrastructure
(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{I$_2$=5}).

@@

\paragraph{Global complexity} Service invocation requires a workflow service
in addition to the science gateway and infrastructure
(\texttt{I$_1$}=3). The architecture involves 6 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$},
\texttt{c$_2$} and \texttt{d} (\texttt{I$_2$}=6).

@@

\paragraph{Global complexity} Task encapsulation requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Global complexity} The pool model requires a workflow pool and an
agent in addition to the science gateway and infrastructure
(\texttt{I$_1$=4}). It involves 7 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$}, \texttt{c$_2$}, \texttt{d}
and \texttt{f} (\texttt{I$_2$}=7).

@@

\paragraph{Global complexity} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.} \note{Marc-e}{I understand both points. However, complexity is
  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}

\paragraph{Robustness} The two workflow services are specific to
workflow execution (\texttt{R$_1$=2}). Interactions \texttt{b}
(counted twice), \texttt{c$_2$} (twice) and \texttt{d} (twice) also
are (\texttt{R$_2$=6}).

@@

\paragraph{Global complexity} Workflow import with service invocation
requires the same components as for service invocation
(\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
for workflow import (\texttt{I$_1$}=7).

@@

  Global complexity (normalized total) &

@@

  are obtained by summing the normalized total scores ($m'$ values) of

>>>>

\definecolor{ao(english)}{rgb}{0.0, 0.5, 0.0}
\newcommand{\todo}[2]{\pdfmargincomment[color=red,author=#1,open=true]{#2}}
\newcommand{\note}[2]{\pdfmargincomment[color=yellow,author=#1,open=true]{#2}}
\newcommand{\closednote}[4]{} % closed notes are hidden. This should obviously be a parameter of \note rather than a separate command.
\newcommand{\answerednote}[4]{\pdfmargincomment[color=green,author=#1-#3,open=true]{#1: #2 \textCR \textCR #3: #4}}
\newcommand{\closedanswerednote}[6]{}

@@

\begin{abstract}
  We investigate software architectures commonly used to integrate
  workflow engines in science gateways. We abstract software
  components and interactions in 6 architectures that we use to
  classify 18 systems. We compare these architectures using metrics
  measuring integration challenge, robustness, extensibility, scalability
  and other specific features. Results show the pros and cons of the
  different architectures: tight integration and task encapsulation are the
  simplest architectures and should therefore be the easiest to
  implement and the most robust ones, most architectures perform the
  same in terms of extensibility, the pool model is the most scalable
  and meta-workflows are only available in nested workflows and
  workflow import. Our analysis provides insights for software
  architects on how to integrate science gateways and workflow
  engines.
\end{abstract}

@@

The remainder of this Section provides background information and
definitions about workflow engines, science gateways and
infrastructures. In Section~\ref{sec:architectures}, we describe 6
architectures in a consistent framework based on the functional
interactions between the main software components involved. In
Section~\ref{sec:evaluation}, we introduce metrics to measure
integration challenge, robustness, extensibility, scalability and other
specific features of the architectures. We use these metrics to
evaluate and compare the 6 architectures. The paper closes on a
discussion and conclusion that highlight architectural recommendations
for science gateway and workflow engine architects.

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration challenge, execution complexity,
extensibility, scalability and other specific features. Criteria break
down to specific metrics where \emph{lower value indicates better
  performance}. For each criterion, a global score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that they can be
summed. The criteria and metrics are explained hereafter.

@@

\cellcolor[HTML]{EEEEEE}\textbf{Integration challenge}& \multicolumn{6}{l}{\cellcolor[HTML]{EEEEEE}}\\

@@

\emph{Integration challenge} is obtained by counting the total number of
interactions and components on the architecture diagram. It breaks
down to the following 2 metrics:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Total number of components (\texttt{I$_1$}), and
\item Total number of interactions (\texttt{I$_2$}).
\end{itemize}
One may wonder whether infrastructure should be counted in
\texttt{I$_1$} since it is usually not developed by the groups who
integrate workflow engines in science gateway. However, integrating an
infrastructure in the architecture usually requires some technical
effort (e.g., account creation, software installation, APIs, etc.), which is
why we keep it in the metric. The two metrics are summed to obtain a
global score that measures the
total number of software pieces to develop, including interactions and
components. It may be used to measure the development effort required
to build the architecture.

\emph{Robustness of workflow execution} measures the likelihood that
workflow execution fails due to errors in the components or with the
interactions in the software architecture. Note that errors coming
from the infrastructure (e.g., unavailable data or terminated tasks)
or workflows (e.g., wrong user input or application errors) are not
covered since they do not stem directly from the software
architecture. Robustness is measured here as a consequence of global
complexity since complex architectures tend to be more prone to
failure. More precisely, robustness is determined as the number of
software components and interactions that are specific to workflow
execution:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Components (\texttt{R$_1$}): number of specific components
  involved in workflow executions. Science
gateway, for instance, is \emph{not} specific to workflow execution
since it is used to authenticate users, add new workflows, transfer
user data, etc.
\item Interactions (\texttt{R$_2$}): number of specific interactions
  involved in workflow executions.
\end{itemize}
These specific components and interactions are the ones represented in
red in the architecture diagrams in
Figure~\ref{fig:architectures}. \texttt{R$_1$} and \texttt{R$_2$} are
summed to obtain the global score for this criterion, which measures
the total number of software pieces that are specifically involved in
the workflow execution.

@@

\paragraph{Integration challenge} This architecture does not require any
component in addition to the science gateway and infrastructure
(\texttt{I$_1$=2}). It involves 5 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$} and \texttt{c$_2$}
(\texttt{I$_2$=5}).

@@

\paragraph{Integration challenge} Service invocation requires a workflow service
in addition to the science gateway and infrastructure
(\texttt{I$_1$}=3). The architecture involves 6 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c$_1$},
\texttt{c$_2$} and \texttt{d} (\texttt{I$_2$}=6).

@@

\paragraph{Integration challenge} Task encapsulation requires only 2
components (\texttt{I$_1$}=2).  It involves 5 interactions:
\texttt{a$_1$}, \texttt{a$_2$}, \texttt{b}, \texttt{c} and
\texttt{e} (\texttt{I$_2$}=5).

@@

\paragraph{Integration challenge} The pool model requires a workflow pool and an
agent in addition to the science gateway and infrastructure
(\texttt{I$_1$=4}). It involves 7 interactions: \texttt{a$_1$},
\texttt{a$_2$}, \texttt{b}, \texttt{c$_1$}, \texttt{c$_2$}, \texttt{d}
and \texttt{f} (\texttt{I$_2$}=7).

@@

\paragraph{Integration challenge} Setting up a nested workflow
architecture with service invocation requires a science gateway, 2
workflow services and 2 infrastructures (\texttt{I$_1$=5}). The
architecture involves 11 interactions (\texttt{I$_2$=11}):
\texttt{a$_1$} (twice), \texttt{a$_2$}, \texttt{b} (twice), \texttt{c$_1$} (twice),
\texttt{c$_2$} (twice) and
\texttt{d} (twice).
\closedanswerednote{Rafael}{I like the idea of what has been discussed in
  the `integration effort', however I think the name may lead to
  misinterpretations. For example, according to Table 2, it requires
  more effort to integrate in a Tight architecture than in a
  service-based architecture. Considering the principles of software
  engineer, it is simpler to integrate/maintain a system using
  decoupled services. Maybe it should be renamed to something related
  to 'dependencies'.}{Tristan}{In our scope I'm not sure that it's
  simpler to integrate a service architecture than a tightly
  integrated one. But I take your point. 'Dependencies' is not very
  suitable either. What about 'development effort'?}{Tristan}{Replaced
  'Integration effort' with 'global complexity', as discussed in
  github issue \#8.}

\paragraph{Robustness} The two workflow services are specific to
workflow execution (\texttt{R$_1$=2}). Interactions \texttt{b}
(counted twice), \texttt{c$_2$} (twice) and \texttt{d} (twice) also
are (\texttt{R$_2$=6}).
\note{Marc-e}{I understand both points. However, complexity is
  addressed in Robustness as well... I liked the idea of development effort. It's a tough call.}

@@

\paragraph{Integration challenge} Workflow import with service invocation
requires the same components as for service invocation
(\texttt{I$_1$}=3). It requires an additional \texttt{g} interaction
for workflow import (\texttt{I$_1$}=7).

@@

  Integration (normalized total)    &

@@

  are obtained by summing the normalized global scores ($m'$ values) of
